---
title: "Perinatal Health Surveillance – United States, 2022"
subtitle: "CDC Natality Data – Smoking, BMI, and Preterm Birth"
author: "Sophie Séguin"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
  html_document:
    toc: true
    toc_float: true
params:
  use_database: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(tidyverse)
library(DBI)
library(RPostgres)
library(knitr)
library(gtsummary)   # for Table 1
library(sandwich)    # robust standard errors
library(lmtest)      # coeftest
library(broom)       # tidy model output
library(forestplot)  # forest plot (optional, ggplot2 can also do it)
```

## Data Source and Methods

The data are from the **CDC Natality Public Use File** for 2022, accessed via the NBER.  
We sampled 500,000 live births (stratified by preterm status) and stored them in a PostgreSQL star schema.

**Definitions:**  
- **Preterm birth**: gestational age < 37 weeks (`gestrec3 == 1`).  
- **Smoking pattern**: derived from cigarettes before pregnancy and during each trimester.  
- **BMI category**: pre‑pregnancy BMI (WHO cutoffs).  
- **Risk ratios**: estimated using modified Poisson regression with robust standard errors.

```{r load_data}
if (params$use_database) {
  con <- dbConnect(
    RPostgres::Postgres(),
    dbname   = "postgres",
    host     = "localhost",
    port     = 5432,
    user     = "postgres",
    password = "1234"        # <<< replace
  )
  
  # Query fact table joined with dimensions
  df <- dbGetQuery(con, "
    SELECT
  m.age_group            AS maternal_age_group,
  m.race_eth,
  m.education,
  m.bmi_cat,
  m.prior_preterm,
  m.diabetes,
  d.c_section,
  i.sex,
  f.preterm_flag,
  f.birth_weight,
  f.apgar5,
  f.lbw_flag            AS lbw,
  f.nicu_flag           AS nicu,
  f.smoking_pattern,
  f.smoke_before,
  f.smoke_tri1,
  f.smoke_tri2,
  f.smoke_tri3
FROM fact_births f
JOIN dim_maternal m ON f.maternal_id = m.maternal_id
JOIN dim_delivery d ON f.delivery_id = d.delivery_id
JOIN dim_infant   i ON f.infant_id   = i.infant_id
  ")
  dbDisconnect(con)
} else {
  # Option 2: Read CSV directly (fallback)
  df <- read_csv("data/processed/natal_cleaned.csv")
}

# Ensure preterm_flag is a factor with 'term' as reference
df <- df %>%
  mutate(
    preterm_flag = factor(preterm_flag, levels = c("term", "preterm")),
    smoking_pattern = factor(smoking_pattern,
      levels = c("never", "quit_early", "quit_late", "started", "persistent")),
    bmi_cat = factor(bmi_cat,
      levels = c("Normal", "Underweight", "Overweight", "Obese")),
    maternal_age_group = factor(maternal_age_group,
      levels = c("20-34", "<20", "35+")),
    race_eth = factor(race_eth),
    education = factor(education,
      levels = c("College grad", "Some college", "High school", "<High school", "Advanced degree")),
    prior_preterm = factor(prior_preterm, levels = c("No", "Yes")),
    diabetes = factor(diabetes, levels = c("No", "Yes")),
    c_section = factor(c_section, levels = c("Vaginal", "C-section")),
    sex = factor(sex, levels = c("Male", "Female"))
  )
```

## Results

### Table 1 – Maternal and Pregnancy Characteristics by Preterm Status

```{r table1}
tbl1 <- df %>%
  select(
    preterm_flag,
    maternal_age_group,
    race_eth,
    education,
    bmi_cat,
    prior_preterm,
    diabetes,
    c_section,
    sex
  ) %>%
  tbl_summary(by = preterm_flag, missing = "ifany", percent = "row") %>%
  add_p() %>%
  modify_header(label = "**Characteristic**")

tbl1
```

### Prevalence of Preterm Birth by Smoking Pattern

```{r prevalence_smoking}
prev_smoke <- df %>%
  filter(!is.na(smoking_pattern)) %>%
  group_by(smoking_pattern) %>%
  summarise(
    n = n(),
    preterm_n = sum(preterm_flag == "preterm"),
    preterm_pct = 100 * mean(preterm_flag == "preterm")
  )

kable(prev_smoke, digits = 1, caption = "Preterm birth prevalence by smoking pattern")
```

### Risk Ratio Analysis

We model preterm birth (preterm vs. term) using modified Poisson regression with robust standard errors to obtain risk ratios.

```{r rrmodel}
# Unadjusted model for smoking pattern
mod1 <- glm((preterm_flag == "preterm") ~ smoking_pattern,
            data = df, family = poisson)
# Robust SE
cov1 <- vcovHC(mod1, type = "HC0")
rr1 <- exp(coef(mod1)[-1])  # vs 'never'
ci1 <- exp(coefci(mod1, vcov = cov1, level = 0.95))[-1, ]
rr_table1 <- data.frame(
  Variable = gsub("smoking_pattern", "", names(rr1)),
  RR = round(rr1, 2),
  CI = sprintf("%.2f–%.2f", ci1[,1], ci1[,2])
)

# Adjusted model: smoking + BMI + age + education + prior preterm + diabetes
mod2 <- glm((preterm_flag == "preterm") ~ smoking_pattern + bmi_cat +
              maternal_age_group + education + prior_preterm + diabetes,
            data = df, family = poisson)
cov2 <- vcovHC(mod2, type = "HC0")
rr2 <- exp(coef(mod2)[-1])
ci2 <- exp(coefci(mod2, vcov = cov2, level = 0.95))[-1, ]

# Prepare forest plot data
plot_data <- data.frame(
  var = names(rr2)[1:length(rr2)],        # adjust for intercept removal
  rr = round(rr2, 2),
  lower = round(ci2[seq(1, nrow(ci2), 1), 1], 2),
  upper = round(ci2[seq(1, nrow(ci2), 1), 2], 2)
) %>% filter(!grepl("Intercept", var)) %>%
  mutate(var = str_remove_all(var, "smoking_pattern|bmi_cat|maternal_age_group|education|prior_preterm|diabetes"))
```

### Adjusted Risk Ratios (Forest Plot)

```{r forestplot, fig.height=6}
ggplot(plot_data, aes(x = rr, y = reorder(var, rr))) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey") +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
  scale_x_log10() +
  labs(x = "Adjusted Risk Ratio (95% CI)", y = "") +
  theme_minimal()
```

## Conclusion

Persistent smoking during pregnancy is strongly associated with preterm birth, even after adjusting for BMI and other confounders. Obesity also increases the risk. This automated report can be regenerated whenever the underlying data update, providing a reproducible BI pipeline.
